Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections at the LArIAT experiment

### 2018-05-31 Chapter-1/1-Theory.tex
s

for the chiral fields $\psi_R$ and $\psi_L$, whose evolution in space and time is coupled through the mass $m$.
If the fermion is massless, the chiral fields decouple and the fermion can be described by a single Weyl spinor with two independent components~\cite{Weyl:10.2307}. Pauli initially rejected the description of a physical particle through a single Weyl spinor because of its implication of parity violation. In fact, since the spatial inversion operator throws $\psi_R \leftrightarrow \psi_L$, parity is conserved only if both chiral components exist at the same time.  For the neutrino introduction in the SM, experiments came in help of the theoretical description.  The constraint of parity conservation weakened after Wu's experiment in 1957 \cite{PhysRev.105.1413}. Additionally,  there was no experimental indication for massive neutrinos, nor evidence of interaction via the neutrino right-handed component.% neutrinos likely interacted only via the left-handed component.

@@

where $c$ e $s$ stand respectively for cosine and sine of the corresponding  mixing angles ($\theta_{12}$, $\theta_{23}$ and $\theta_{13}$), $\delta$ is the Dirac CP violation phase, $\alpha_1$ and $\alpha_{2}$  are the eventual Majorana CP violation phases.  Experimental results on neutrino oscillations are generally reported in terms of the mixing angles and of the squared mass splitting $\Delta m^2_{ab} = m^2_{a} - m^2_{b}$, where $a$ and $b$ represent the mass eigenstates. A summary of the current status of experimental results, albeit partial, is given in table \ref{tab:nuosc}.

@@

Historically, neutrino interactions with the nucleus in the GeV region are divided into three categories into three categories whose contributions change as a function of increasing neutrino energy:: quasi elastic (QE), resonant (RES), and deep inelastic (DIS) scattering. All current and forthcoming oscillation experiments on neutrino beams live in the 0.1-10 GeV transition region, which encompasses the energy where the QE neutrino-nucleus interaction transitions into RES and then into DIS. 
For scattering off free nucleons, neutrino and antineutrino QE charge current scattering refers to the process $\nu_l n \rightarrow l^- p$ and $\bar\nu_l p \rightarrow l^+ n$ where a charged lepton and single nucleon are ejected in the elastic interaction. %of a neutrino (or antineutrino) with a nucleon in the target material.  
 Resonant scattering refers to an inelastic collision producing a nucleon excited state ($\Delta$,N*) -- the resonance -- which then quickly decays, most often to a nucleon and single-pion final state. DIS refers to the head-on collision between the neutrino and a parton inside the nucleon, producing hadronization and subsequent  abundant production of mesons and nucleons.  In addition to such interactions between the neutrino and a single component of the nucleus, neutrinos can also interact with the nucleus as a whole, albeit more rarely, a well documented process called coherent meson production scattering \cite{Vilain1993}; the signature of such process is the production of a distinctly forward-scattered single meson final state, most often a pion.  This simple picture of neutrino interactions works rather well for scattering 
off of light nuclear targets, such as the H$_2$ and D$_2$ of bubble chamber experiments \cite{RevModPhys.84.1307}, but the complexity of the nuclear structure for heavier nuclei such as argon complicates this model.

@@

A renewed interest for neutrino cross section measurements surged in recent years, along with a lively discussion on the data reporting; the historical method of reporting the neutrino cross section as a function of the neutrino energy or momentum transferred shakes under the weight of its dependency on the chosen nuclear model.   On one hand, correcting for nuclear effects in neutrino interaction can introduce unwanted sources of uncertainty and model dependency especially due to the mis-modeling of the meson interactions. On the other, avoiding this correction makes a comparison between neutrino interactions on different target nuclei extremely difficult.

@@

\textbf{SUSY GUTs.} Supersymmetry theories allow for another family of GUTs. In SUSY, every fundamental particle in the SM has a ``superpartner", identical in each quantum number except for the spin-statistics: the fermion supersymmetric partners are bosons and vice versa. Collider experiments (mainly LHC)  constrain the mass of the supersymmetric partners to be very heavy\cite{LHC}.
The SU(5), SU(10) groups with a SUSY twist are the basic groups for SUSY GUTs. From the phenomenology point of view, SUSY models tend to push the proton decay life time higher by a factor of four, they solve the ``hierarchy problem", and they also predict new channels for the proton decay. In particular they predict the presence of kaons in the final product, with a dominant mode of $p\rightarrow K^+ \bar\nu$. %, since particles from the first generation (e.g. $u$ and $d$) need to decay into second and third generation particles (such as $s$). 
Predictions on the proton decay lifetime depend on the chosen SUSY model; again, some of the predictions are not excluded by the experiments \cite{PhysRevD.55.6986, PhysRevD.67.075007,Lopez:1991eca}.

@@

\begin{itemize}
\item[-] the assessment of the existence of right-handed sterile neutrinos via the study of accelerator neutrinos on a short baseline (SBN), 
\item[-] the determination of the sign of $\Delta m^2_{13}$ (or $\Delta m^2_{23}$), i.e., the neutrino mass hierarchy via the study of accelerator neutrinos on a long baseline (DUNE),
\item[-] the determination of the octant, i.e.  whether $\theta_{23}$ is maximal, via the study of accelerator neutrinos on a long baseline (DUNE),
\item[-] the determination the status of CP symmetry in the lepton sector, via the study of accelerator neutrinos on a long baseline (DUNE),
\item[-] the search for observables predicted by GUTs, such as proton decay via the study of non accelerator physics in massive underground detectors (DUNE).
\end{itemize}

@@

The key elements of a neutrino event generators for resonance and DIS events are the nuclear model and the hadron treatment (both production and transportation).  We illustrate here the conceptual basis of the GENIE Neutrino Generator~\cite{Andreopoulos:2009rq} as an example, since GENIE is one the most popular event generators for liquid argon experiments.
For example, the nuclear model used by GENIE for all processes is a Relativistic Fermi Gas (RFG) model modified to incorporate nucleon-nucleon correlations \cite{PhysRevD.24.1400}. This means that the initial momentum and binding energy of the struck nucleon is determined by assuming nucleons inside the nucleus are quasi-free, acting independently in the mean field of the nucleus. For A $ >$ 20 such as argon, the 2-parameter Woods-Saxon shell model for density function is used.  The GENIE module INTRANUKE \cite{Juszczak2006} is used to simulate final-state interactions (FSI) which model  hadron re-interactions inside the nucleus. This module places the outgoing particles in the nucleus and propagates them using the ``hA model".  In the INRANUKE hA model, hadrons can undergo at most one FSI per event. When possible, external hadron-nucleus scattering data are used to tune INTRANUKE. Since no data is available for Argon, GENIE uses an interpolation of data from heavier and lighter nuclei for the pion-argon cross section leading to large (10?s of \%) resultant uncertainties in the INTRANUKE module.

@@

LArIAT tiny active volume makes it impossible for the experiment to place competitive limits on nucleon decay searches.  However,  LArIAT provides excellent data to characterize kaons in liquid argon for the ``LAr golden mode", $p \rightarrow K^+ \bar{\nu}$.  The result of these studies will affect future proton decay searches in LArTPCs.  Previous work has been done to assess the potential identification efficiency for different decay modes in a LArTPC \cite{Bueno2007}, but, as the time of this  writing, no study of kaon selection efficiency in LArTPCs has been performed on data. 
The K$^+$-Ar interaction cross section has never been measured before and can affect the possibility of detecting and measuring kaons when produced in a proton decay event. 
Kaon interactions with argon can distort the kaon energy spectrum as well as change the topology of single kaon events. In a LArTPC, non-interacting kaons appear as straight tracks with a high ionization depositions at the end (Bragg peak). The topology of interacting kaons can be quite different. In case of elastic scattering, a distinct kink will be present in the track. In case of inelastic scattering the Bragg peak will not be present and additional tracks will populate the event.
Performing the total hadronic K$^+$-Ar cross section measurement on data serves the double purpose of identifying the rate of ``unusual" topologies (kinks and additional tracks) and of developing tools for kaon tracking in LAr.

>>>>

for the chiral fields $\psi_R$ and $\psi_L$, whose evolution in space and time is coupled through the mass $m$.
If the fermion is massless, the chiral fields decouple and the fermion can be described by a single Weyl spinor with two independent components~\cite{Weyl:10.2307}. Pauli initially rejected the description of a physical particle through a single Weyl spinor because of its implication of parity violation. In fact, since the spatial inversion operator throws $\psi_R \leftrightarrow \psi_L$, parity is conserved only if both chiral components exist at the same time.  For the neutrino introduction in the SM, experiments were critical to a more accurate theoretical description.  The constraint of parity conservation weakened after Wu's experiment in 1957 \cite{PhysRev.105.1413}. Additionally,  there was no experimental indication for massive neutrinos, nor evidence of interaction via the neutrino right-handed component.% neutrinos likely interacted only via the left-handed component.

@@

where $c$ and $s$ stand respectively for cosine and sine of the corresponding  mixing angles ($\theta_{12}$, $\theta_{23}$ and $\theta_{13}$), $\delta$ is the Dirac CP violation phase, $\alpha_1$ and $\alpha_{2}$  are the eventual Majorana CP violation phases.  Experimental results on neutrino oscillations are generally reported in terms of the mixing angles and of the squared mass splitting $\Delta m^2_{ab} = m^2_{a} - m^2_{b}$, where $a$ and $b$ represent the mass eigenstates. A summary of the current status of experimental results is given in table \ref{tab:nuosc}.

@@

Historically, neutrino interactions with the nucleus in the GeV region are divided into three categories  whose contributions change as a function of increasing neutrino energy:: quasi elastic (QE), resonant (RES), and deep inelastic (DIS) scattering. All current and near future oscillation experiments on neutrino beams live in the 0.1-10 GeV transition region, which encompasses the energy where the QE neutrino-nucleus interaction transitions into RES and then into DIS. 
For scattering off free nucleons, neutrino and antineutrino QE charge current scattering refers to the process $\nu_l n \rightarrow l^- p$ and $\bar\nu_l p \rightarrow l^+ n$ where a charged lepton and single nucleon are ejected in the elastic interaction. %of a neutrino (or antineutrino) with a nucleon in the target material.  
 Resonant scattering refers to an inelastic collision producing a nucleon excited state ($\Delta$,N*) -- the resonance -- which then quickly decays, most often to a nucleon and single-pion final state. DIS refers to the head-on collision between the neutrino and a parton inside the nucleon, producing hadronization and subsequent  abundant production of mesons and nucleons.  In addition to such interactions between the neutrino and a single component of the nucleus, neutrinos can also interact with the nucleus as a whole, albeit more rarely, a well documented process called coherent meson production scattering \cite{Vilain1993}; the signature of such process is the production of a distinctly forward-scattered single meson final state, most often a pion.  This simple picture of neutrino interactions works rather well for scattering 
off of light nuclear targets, such as the H$_2$ and D$_2$ of bubble chamber experiments \cite{RevModPhys.84.1307}, but the complexity of the nuclear structure for heavier nuclei such as argon complicates this model.

@@

There has been a renewed interest in neutrino cross section measurements in recent years, along with a lively discussion on the data reporting; the historical method of reporting the neutrino cross section as a function of the neutrino energy or momentum transferred shakes under the weight of its dependency on the chosen nuclear model.   On one hand, correcting for nuclear effects in neutrino interaction can introduce unwanted sources of uncertainty and model dependency especially due to the mis-modeling of the meson interactions. On the other, avoiding this correction makes a comparison between neutrino interactions on different target nuclei extremely difficult.

@@

\textbf{SUSY GUTs.} Supersymmetry theories allow for another family of GUTs. In SUSY, every fundamental particle in the SM has a ``superpartner", identical in each quantum number except for the spin-statistics: the fermion supersymmetric partners are bosons and vice versa. Collider experiments (mainly LHC)  constrain the mass of the supersymmetric partners to be very heavy.
The SU(5), SU(10) groups with a SUSY twist are the basic groups for SUSY GUTs. From the phenomenology point of view, SUSY models tend to push the proton decay life time higher by a factor of four, they solve the ``hierarchy problem", and they also predict new channels for the proton decay. In particular they predict the presence of kaons in the final product, with a dominant mode of $p\rightarrow K^+ \bar\nu$. %, since particles from the first generation (e.g. $u$ and $d$) need to decay into second and third generation particles (such as $s$). 
Predictions on the proton decay lifetime depend on the chosen SUSY model; again, some of the predictions are not excluded by the experiments \cite{PhysRevD.55.6986, PhysRevD.67.075007,Lopez:1991eca}.

@@

\begin{itemize}
\item[-] the assessment of the existence of sterile neutrinos via the study of accelerator neutrinos on a short baseline (SBN), 
\item[-] the determination of the sign of $\Delta m^2_{13}$ (or $\Delta m^2_{23}$), i.e., the neutrino mass hierarchy via the study of accelerator neutrinos on a long baseline (DUNE),
\item[-] the determination of the octant, i.e.  whether $\theta_{23}$ is maximal, via the study of accelerator neutrinos on a long baseline (DUNE),
\item[-] the determination the status of CP symmetry in the lepton sector, via the study of accelerator neutrinos on a long baseline (DUNE),
\item[-] the search for observables predicted by GUTs, such as proton decay via the study of non accelerator physics in massive underground detectors (DUNE).
\end{itemize}

@@

The key elements of a neutrino event generators for resonance and DIS events are the nuclear model and the hadron treatment (both production and transportation).  We illustrate here the conceptual basis of the GENIE Neutrino Generator~\cite{Andreopoulos:2009rq} as an example, since GENIE is one the most popular event generators for liquid argon experiments.
For example, the nuclear model used by GENIE for all processes is a Relativistic Fermi Gas (RFG) model modified to incorporate nucleon-nucleon correlations \cite{PhysRevD.24.1400}. This means that the initial momentum and binding energy of the struck nucleon is determined by assuming nucleons inside the nucleus are quasi-free, acting independently in the mean field of the nucleus. For A $ >$ 20 such as argon, the 2-parameter Woods-Saxon shell model for density function is used.  The GENIE module INTRANUKE \cite{Juszczak2006} is used to simulate final-state interactions (FSI) which model  hadron re-interactions inside the nucleus. This module places the outgoing particles in the nucleus and propagates them using the ``hA model".  In the INRANUKE hA model, hadrons can undergo at most one FSI per event. When possible, external hadron-nucleus scattering data are used to tune INTRANUKE. Since no data is available for Argon, GENIE uses an interpolation of data from heavier and lighter nuclei for the pion-argon cross section leading to resultant uncertainties in the INTRANUKE module of the order of 10-20\%.

@@

LArIAT's tiny active volume makes it impossible for the experiment to place competitive limits on nucleon decay searches.  However,  LArIAT provides excellent data to characterize kaons in liquid argon for the ``LAr golden mode", $p \rightarrow K^+ \bar{\nu}$.  The result of these studies will affect future proton decay searches in LArTPCs.  Previous work has been done to assess the potential identification efficiency for different decay modes in a LArTPC \cite{Bueno2007}, but, as the time of this  writing, no study of kaon selection efficiency in LArTPCs has been performed on data. 
The K$^+$-Ar interaction cross section has never been measured before and can affect the possibility of detecting and measuring kaons when produced in a proton decay event. 
Kaon interactions with argon can distort the kaon energy spectrum as well as change the topology of single kaon events. In a LArTPC, non-interacting kaons appear as straight tracks with a high ionization depositions at the end (Bragg peak). The topology of interacting kaons can be quite different. In case of elastic scattering, a distinct kink will be present in the track. In case of inelastic scattering the Bragg peak will not be present and additional tracks will populate the event.
Performing the total hadronic K$^+$-Ar cross section measurement on data serves the double purpose of identifying the rate of ``unusual" topologies (kinks and additional tracks) and of developing tools for kaon tracking in LAr.
